---
title: "Modellierung und Simulation"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 6.1: Discrete time markov chains
```{r,echo=FALSE}
library(ggplot2)
```
We define a function for drawing the transition diagram
```{r}
transitionDiagram <- function(df,curv=1,loopSize=0.5,arrowSize=0.025,pointCol='black',transCol='grey',selfCol='grey',selfAngle=90){
  ##
  # creates a transition Diagram from a dataframe
  ##

  numStates <- length(unique(c(df$From,df$To)))

  df1 <- df[df['From']==df['To'],]  # all self-directed (loop) state transition
  nv1 <- rep(0,dim(df1)[1])
  ov1 <- rep(loopSize,dim(df1)[1])
  df2 <- df[df['From']!=df['To'],]  # all non-self transitions
  nv2 <- rep(0,dim(df2)[1])
  ov2 <- rep(loopSize,dim(df2)[1])
  
  # draw state transitions
  diagram <- ggplot(data = df2,aes(x=From,y=ov2)) + 
    geom_point(aes(x=From,y=nv2),size=3,pch=16,colour=pointCol) +
    geom_point(aes(x=To,y=nv2),size=3,pch=16,colour=pointCol) +
    geom_text(aes(x=From,y=nv2),label=paste0("        ",df2$From)) +
    geom_text(aes(x=To,y=nv2),label=paste0("        ",df2$To)) +
    geom_curve(aes(x=From,y=nv2,xend=To,yend=nv2),curvature = curv, ncp=1000, color=transCol,
             arrow = arrow(length=unit(arrowSize,"npc"),type = "closed"))  # draw transition to next other status
    
  # fit the overall plot size to the loopSize
  if(loopSize>1){
    diagram <- diagram + xlim(-loopSize+1,numStates+loopSize-1) + ylim(-loopSize,loopSize)
  } else {
    diagram <- diagram + xlim(-1, numStates) + ylim(-1, 1)
  }
    
  # if there are self loops draw them here (as 2 seperate arrow for 1 loop)
  if(sum(df['From']==df['To']) > 0){ 
    diagram <- diagram + 
      geom_curve(data = df1, aes(x = From, y = nv1, xend = To, yend = ov1),curvature = curv, ncp=1000, color=selfCol, angle=selfAngle) +  # draw transition from same status
      geom_curve(data = df1, aes(x = From, y = ov1, xend = To, yend = nv1),curvature = curv, ncp=1000, color=selfCol, angle=selfAngle, arrow = arrow(length=unit(arrowSize,"npc"),type = "closed")) # draw transition to same status
  }
  
  blank <- theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
  
  return(diagram + blank)
}
```
### 1.
After that we can get the information from the exercise example and build from it the respective graph. In addition we draw the adjacency matrix/state transition matrix.

TODO: shall we draw the probabilities on top of the arcs?
```{r}
adjVector <- c(0.1,0.2,0.3,0.4,0,0,1,0,0,0,0,1,1,0,0,0)
stateTransitionMatrix <- matrix(adjVector,nrow=4,ncol=4,dimnames=list(c(0,1,2,3),c(0,1,2,3)),byrow=TRUE)
stateTransitionMatrix

States <- c(0,1,2,3)
graph <- data.frame(From=rep(States,each=4),To=rep(States,4),Probability=adjVector)
exGraph= graph[graph$Probability!=0,]

transitionDiagram(df=exGraph,curv=0.5)
transitionDiagram(df=exGraph)
```

### 2.
Again we start with defining a function, this time for calculation the transition vectors and printing them.
```{r}
stateDistrVecs <- function(x_0,n_max=10,transMat,X_start=0){
  sdv <- x_0  # intial start vector
  tm <- transMat
  sdv <- rbind(sdv,tm[X_start+1,])  # first step
  for(n in 2:n_max){  # every following step
    tm <- tm %*% transMat
    sdv <- rbind(sdv,tm[X_start+1,])
  }
  rownames(sdv) <- c(paste0("x_",0:n_max))
  return(sdv)
}

```
We start with `x_0 = (1,0,0,0)`, as our initial state is chosen **0**.
```{r}
x_0 <- c(1,0,0,0)
stateDistrVecs(x_0=x_0,transMat=stateTransitionMatrix)
```

### 3.
The *stationary state distribution* can be computed by going against `n_max = math.inf` when powering the state transition matrix. 
```{r}
getStationaryStateDistr <- function(transMat,n_max){
  tm <- stateTransitionMatrix
  for(n in 0:n_max){
    tm <- tm %*% stateTransitionMatrix
  }
  return(tm[1,])
}
```
As this will quickly approach the \"long run\" solution, we take `n_max = 1000` here.

Us starting at state **0** is assumed to be still given.
```{r}
getStationaryStateDistr(transMat=stateTransitionMatrix,n_max=1000)
```

### 4./5.
Same procedure as above for changed probabilities in the state transitions. `p_{01} = 0.5, p_{03} = 0.5`.
```{r}
new_0_probs <- c(0,0.5,0,0.5)
stateTransitionMatrix[1,] <- new_0_probs
stateTransitionMatrix

adjVector[1:4] <- new_0_probs
graph <- data.frame(From=rep(States,each=4),To=rep(States,4),Probability=adjVector)
exGraph= graph[graph$Probability!=0,]

transitionDiagram(df=exGraph)
```

### 6.
```{r}
stateDistrVecs(x_0=x_0,transMat=stateTransitionMatrix)
```


### 7.
We do observe an oscillation of the state distributions, e.g. consecutive state distributions do not converge.
The oscillation period is `p = 2`, as we can deduce easily from the example vectors **x_0** to **x_10**.

### 8.
As we got oscillating state distributions and `p = 2`, we got 2 different possibilities for the different states in the long run. 

*(see 8.10 Example 3 again)*
```{r}
k <- 0
getStationaryStateDistr(transMat=stateTransitionMatrix,n_max=1000+k)

k <- 1
getStationaryStateDistr(transMat=stateTransitionMatrix,n_max=1000+k)
```

### 9.
It would **not** have been sufficient to remove only one transition.

This is due to the fact that with **4** different states and **3** transitions from state **0** we cannot reach an oscillation point with `p = 2`. This would be the only possible value for a period in a Markov Chain with **4** states.

## Problem 6.2: Repeated coin flipping
### 1.
State space: `X = {0,1,2,3,4,5,6}: number of points reached`.

Set of terminating states: `X_s = {5,6}`.

Factor space: `Y = {1,2}: points per coinflip`

### 2.
```{r}
te <- NULL
probMat <- matrix(c(rep(c(0,0.5,0.5,0,0,0,0,0),5),c(rep(0,9))),nrow=7,ncol=7,byrow=TRUE)
colnames(probMat) <- 0:6
rownames(probMat) <- 0:6
probMat

for(x in 0:4){
  for(y in (x+1):(x+2)){
    te <- rbind(te,paste0("f(",x,",",y,") = 0.5"))
  }
}
colnames(te) <- "equation"
rownames(te) <- 1:10
te
```

### 3.
```{r}
df <- which(probMat>0,arr.ind=TRUE)
df <- as.data.frame(df)
df$V3 <- rep(0.5,dim(df)[1])
colnames(df) <- c("From","To","Probabilities")
rownames(df) <- 1:dim(df)[1]
df$From <- df$From - 1
df$To <- df$To - 1
df

transitionDiagram(df)
```

### 4.
Here we extract the distributions (presented as a data frame) for winning the game.

From this we can easily derive the answers to questions *a)*, *b)* and *c)*.
```{r}
coinflipDistr <- stateDistrVecs(x_0=c(1,0,0,0,0,0,0),n_max=7,transMat=probMat)
#coinflipDistr

distributionEnd <- coinflipDistr[,6:7]
distributionEnd <- distributionEnd[-1,]
distributionEnd
```

